\documentclass[10pt]{article}

% Be sure to use PDF Latex
\pdfoutput=1

% links
\usepackage[bookmarks,bookmarksdepth=2, colorlinks=true, linkcolor=blue,citecolor=red, urlcolor=blue]{hyperref}


\usepackage{fullpage}

\usepackage[latin1]{inputenc}
\usepackage{mystyle}
\usepackage{wrapfig}
\usepackage{notations_ot} % for OT chapter
\usepackage{tikz}

\newcommand{\dims}{d}


\graphicspath{{./figures/}}


\newcommand{\myauthor}[3]{#1\protect\footnote{#2, \protect\url{#3}}}


\title{Course notes on\\ Computational Optimal Transport} 

\author{%
\begin{tabular}{c}
	Gabriel Peyr{\'e} \\ CNRS \& DMA \\
	 \'Ecole Normale Sup\'erieure \\
	 \url{gabriel.peyre@ens.fr}\\
	 \url{https://mathematical-tours.github.io}\\
	 \url{www.numerical-tours.com}
\end{tabular}
}


\date{\today}

%%

\begin{document}

\maketitle

\begin{abstract}
These course notes present the foundational mathematics concepts relevant to computational optimal transport. They delve into various formulations and techniques, starting with Monge's formulation, illustrated through applications to one-dimensional optimal transport and distributions involving Gaussians. The notes then explore Kantorovich's formulation, a more general approach that addresses limitations in Monge's method and applies to a wider range of problems.
%
Additionally, the course covers the concept of the Wasserstein distance, a key metric in optimal transport theory that quantifies the cost of transporting one distribution to another.
%
The notes also introduce entropic regularization, an important technique used to smooth optimization problems and make them more tractable computationally. This approach facilitates the numerical solution of optimal transport problems by adding an entropy term to the objective function, leading to faster and more stable algorithms.
%
Finally, the dual formulation of optimal transport problems is presented. This perspective allows for deriving alternative algorithms and insights into the structure of optimal transport solutions. By exploring the dual problem, students can understand the relationships between different formulations and how they can be exploited to solve transport problems more efficiently.
%
Overall, these course notes provide a comprehensive introduction to the mathematics of computational optimal transport, offering insights into both theoretical foundations and practical applications.
\end{abstract}

% \tableofcontents

\input{sections/matching}
\input{sections/monge}
\input{sections/kantorovitch}
\input{sections/sinkhorn}
\input{sections/dual}
\input{sections/semidiscr-w1}
\input{sections/dual-norms}
\input{sections/sinkhorn-div}
%\input{sections/barycenters}
%\input{sections/estimation}
\input{sections/grad-flows}
%\input{sections/extensions}



% \nocite{*}

\bibliographystyle{plain}
\bibliography{all}

\end{document}
