% ------------------------------------------------- -----
% Table of ratings
% ------------------------------------------------- -----
\listofnotationsstyle{}
\chapter *{Table of notations}
\thispagestyle{empty}
%% set up for supertabular env
\tablefirsthead{\hline Notation & Meaning & Page \\\hline}
\tablehead{\hline Notation & Meaning & Page \\\hline}
\tabletail{\hline}
\tablelasttail{\hline}
%% the supertabular
\begin{center}
\begin{supertabular}{| c | p{9cm} | c |}
$ \wh{G} $ & Dual of a group. & \pageref{notation-0} \\[1mm]
$ \chi $ & Character of a group. & \pageref{notation-1} \\[1mm]
$ \UU_n $ & Set of \ordin{n}{th} roots of the unit. & \pageref{notation-2} \\[1mm]
$ \imath $ & Complex number such as $ \imath^2 = -1 $. & \pageref{notation-3} \\[1mm]
$ | z |, \; \ol{z} $ & Modulus and conjugate of a complex number. & \pageref{notation-4} \\[1mm]
$ \realp (z), \; \imagp (z) $ & Real and imaginary parts. & \pageref{notation-5} \\[1mm]
$ |G| $ & Order of a group $ G $. & \pageref{notation-6} \\[1mm]
$ \CC [G] $ & Spaces of functions from $ G $ in $ \CC $. & \pageref{notation-7} \\[1mm]
$ a \eqdef b $ & Equal by definition. & \pageref{notation-8} \\[1mm]
$ \dotp{f}{g} $ & Hermitian product of two functions. & \pageref{notation-9} \\[1mm]
$ \norm{f}_2 $ & Norm of a function. & \pageref{notation-10} \\[1mm]
$ L^2 (\RR) $ & Integrable square functions. & \pageref{notation-11} \\[1mm]
$ \omega $ & Primitive root of the unit. & \pageref{notation-12} \\[1mm]
$ G \simeq G'$ & Isomorphism (of group, etc). & \pageref{notation-13} \\[1mm]
$ \delta_p^q $ & Kroneker symbol. & \pageref{notation-14} \\[1mm]
$ [G: H] $ & Subscript of $ H $ in $ G $. & \pageref{notation-15} \\[1mm]
$ \whwh{G} $ & Bidual of the group $ G $. & \pageref{notation-16} \\[1mm]
$ \mathfrak{S}_n $ & Symmetric group. & \pageref{notation-17} \\[1mm]
$ D (G) $ & Derivative group of $ G $. & \pageref{notation-18} \\[1mm]
$ [x, \, y] $ & $ x $ and $ y $ switch. & \pageref{notation-19} \\[1mm]
$ \mathfrak{A}_n $ & Group of even permutations. & \pageref{notation-20} \\[1mm]
$ c_f(\chi) $ & Fourier coefficient. & \pageref{notation-21} \\[1mm]
$ \Ff(f) $ & Fourier transform of a function $ f $. & \pageref{notation-22} \\[1mm]
$ f * g $ & Convolution product of $ f $ and $ g $. & \pageref{notation-23} \\[1mm]
$ \text{p.p.t.} $ & For almost everything. & \pageref{notation-24} \\[1mm]
$ \legsymb{n}{p} $ & Legendre symbol. & \pageref{notation-25} \\[1mm]
$ \Tr_{K / k} $ & Trace a finite field $ K $ over $ k $. & \pageref{notation-26} \\[1mm]
$ \psi_1 $ & Canonical character. & \pageref{notation-27} \\[1mm]
$ \eta $ & Quadratic character. & \pageref{notation-28} \\[1mm]
$ G (\chi, \, \psi) $ & Gaussian sum. & \pageref{notation-29} \\[1mm]
$ \Ff_{\text{mul}} (f) $ & Multiplicative Fourier transform. & \pageref{notation-30} \\[1mm]
$ \Ff_{\text{add}} (f) $ & Additive Fourier transform. & \pageref{notation-31} \\[1mm]
$ W_{2^k} $ & Walsh matrices. & \pageref{notation-32} \\[1mm]
$ \Ww_k (f) $ & Walsh transform of $ f $. & \pageref{notation-33} \\[1mm]
$ E^* $ & Dual of a vector space $ E $. & \pageref{notation-34} \\[1mm]
$ A^{\bot} $ & Orthogonal subspace of a part $ A $ of $ E $. & \pageref{notation-35} \\[1mm]
$ F^{0} $ & Orthogonal subspace of a part $ F $ of $ E^* $. & \pageref{notation-36} \\[1mm]
$ H^{\sharp} $ & Orthogonal of a subgroup. & \pageref{notation-37} \\[1mm]
$ d (x, \, y) $ & Hamming distance between two vectors of $ \FF_q^k $. & \pageref{notation-38} \\[1mm]
$ w (z) $ & Weight of a vector of $ \FF_q^k $. & \pageref{notation-39} \\[1mm]
$ A_H (X, \, Y) $ & Enumerator polynomial of a subspace $ H $ of $ (\FF_2)^k $. & \pageref{notation-40} \\[1mm]
$ \Pi_s $ & Dirac comb. & \pageref{notation-41} \\[1mm]
$ \{f [k]\}_{k = 0}^n $ & Finite vector of size $ N $. & \pageref{notation-42} \\[1mm]
$ \wh{f}[k] $ & Discrete Fourier coefficient. & \pageref{notation-43} \\[1mm]
$ f^0, \; f^1 $ & Even and odd parts of a sample $ f $. & \pageref{notation-44} \\[1mm]
$ \Ss_N^x $ & Operator $ \Ss $. & \pageref{notation-45} \\[1mm]
$ f_g, \; f_d $ & Left and right parts of a sample $ f $. & \pageref{notation-46} \\[1mm]
$ M^*, \; \transp{M} $ & Adjunct and transposed matrices. & \pageref{notation-47} \\[1mm]
$ f^\sharp $ & Symmetrized function. & \pageref{notation-48} \\[1mm]
$ f_{s}, \; f_{a} $ & Symmetric and anti-symmetric parts & \pageref{notation-49} \\[1mm]
$ M_n (k) $ & Matrices of size n. & \pageref{notation-50} \\[1mm]
$ GL_n (k) $ & Invertible matrices of size n. & \pageref{notation-51} \\[1mm]
$ \wh{\RR} $ & Dual of $ \RR $. & \pageref{notation-52} \\[1mm]
$ \Phi^g $ & Linear filter. & \pageref{notation-53} \\[1mm]
$ \ell^1 (\ZZ) $ & Absolutely summable suites. & \pageref{notation-54} \\[1mm]
$ \Corr (f, \, g) $ & Correlation of $ f $ and $ g $. & \pageref{notation-55} \\[1mm]
$ \ol{\Corr} (f, \, g) $ & Normalized correlation of $ f $ and $ g $. & \pageref{notation-56} \\[1mm]
$ \beta^n $ & B-spline function of order $ n $. & \pageref{notation-57} \\[1mm]
$ \case (x) $ & Hartley's kernel. & \pageref{notation-58} \\[1mm]
$ \Hh (f) $ & Hartley transform. & \pageref{notation-59} \\[1mm]
$ \Zz (f) $ & Transformed to Z. & \pageref{notation-60} \\[1mm]
$ \Phi_a^b $ & Recursive filter. & \pageref{notation-61} \\[1mm]
$ \Ll (f) $ & Laplace transform. & \pageref{notation-62} \\[1mm]
$ \Gg_z (f) $ & Vector Z transform. & \pageref{notation-63} \\[1mm]
$ \Phi_n $ & Cyclotomic polynomial. & \pageref{notation-64} \\[1mm]
$ \text{Fer}_n $ & \ordin{n}{th} Fermat number. & \pageref{notation-65} \\[1mm]
$ w (x) $ & Weight of a word. & \pageref{notation-66} \\[1mm]
$ d (x, \, y) $ & Hamming distance. & \pageref{notation-67} \\[1mm]
$ \# E $ & Cardinal of a set. & \pageref{notation-68} \\[1mm]
$ \Cc^\bot $ & Dual of a code. & \pageref{notation-69} \\[1mm]
$ GL (V) $ & Linear group of a space $ V $. & \pageref{notation-70} \\[1mm]
$ \delta_g $ & Base function of $ \CC [G] $. & \pageref{notation-71} \\[1mm]
$ K [G] $ & Algebra of a group. & \pageref{notation-72} \\[1mm]
$ \rho_{V \oplus W} $ & Sum representation. & \pageref{notation-73} \\[1mm]
$ V \oplus W $ & Direct sum of subspaces. & \pageref{notation-74} \\[1mm]
$ \Ll (V, \, W) $ & Space of morphisms from $ V $ into $ W $. & \pageref{notation-75} \\[1mm]
$ \rho_{\Ll (V, \, W)} $ & Representation of morphisms. & \pageref{notation-76} \\[1mm]
$ V^* $ & Dual of a vector space $ V $. & \pageref{notation-77} \\[1mm]
$ \rho_{V^*} $ & Dual representation. & \pageref{notation-78} \\[1mm]
$ \transp{\varphi} $ & Application transposed. & \pageref{notation-79} \\[1mm]
$ V^G $ & Invariant under-representation. & \pageref{notation-80} \\[1mm]
$ \Hom_G (V, \, W) $ & Interleaving operators. & \pageref{notation-81} \\[1mm]
$ \Ker (f) $ & Kernel of $ f $. & \pageref{notation-82} \\[1mm]
$ \Im (f) $ & Image of $ f $. & \pageref{notation-83} \\[1mm]
$ \dim_{K} (V) $ & Dimension of a vector space. & \pageref{notation-84} \\[1mm]
$ Z (G) $ & Center of $ G $. & \pageref{notation-85} \\[1mm]
$ R_G $ & Reynolds operator. & \pageref{notation-86} \\[1mm]
$ \tr (f) $ & Trace of endomorphism $ f $. & \pageref{notation-87} \\[1mm]
$ \wt{f} $ & Averaged application. & \pageref{notation-88} \\[1mm]
$ \chi_{\rho} $ & Character of a representation $ \rho $. & \pageref{notation-89} \\[1mm]
$ V_i^{\oplus a_i} $ & Direct sum multiple. & \pageref{notation-90} \\[1mm]
$ \end (V) $ & Endomorphisms of $ V $. & \pageref{notation-91} \\[1mm]
$ \pi f(\rho) $ & Fourier transform from $ f $ to $ \rho $. & \pageref{notation-92} \\[1mm]
$ \CC [G]^G $ & Space of central functions. & \pageref{notation-93} \\[1mm]
$ c_{f} (\rho_i) $ & Fourier coefficient. & \pageref{notation-94} \\[1mm]
$ \Theta (G) $ & Determinant of a group. & \pageref{notation-95} \\[1mm]
$ \Theta_\rho (G) $ & Determinant of a group in a representation. & \pageref{notation-96} \\[1mm]
$ c_{f} (k, \, i, \, j) $ & Generalized Fourier coefficients. & \pageref{notation-97} \\[1mm]
\end{supertabular}
\end{center}
% flush page must be blank 
\clearpage{\pagestyle{empty}\cleardoublepage}
